---
title: "Data wrangling - Deficit as % of GDP"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library (tidyverse)
library (plotly)
library(zoo) # moving averages 
```

## Preparing data

Uploading data 
```{r reading, message=FALSE}
df_deficit_pct <- read_csv("data/deficit_pct_gdp_OECD.csv")
```
Creating a wide-form dataset with columns for each year
There are 23 rows with flag E (estimate/preliminary data) and 1 B (break in series) that were put in table df_e_or_b
```{r}
df_deficit_pct %>% count(`Flag Codes`)
```

```{r}
df_deficit_pct %>% filter(!is.na(`Flag Codes`))
df_wide_deficit <- df_deficit_pct %>% 
  select(-`Flag Codes`) %>% #this removes the variable Flag Codes to make the wide dataset cleaner
  pivot_wider(names_from = TIME, values_from = deficit_pc)
```
Creating a comparators variable
```{r} 
comparators <- c("DNK","IRL","NZL","NOR","SWE")
    df_deficit_pct <- df_deficit_pct %>%
    mutate(in_comparators=if_else(LOCATION%in%comparators,TRUE,FALSE)) 
```

Creating rolling averages over 5, 7 and 10 years
```{r}
  df_deficit_pct <- df_deficit_pct %>% #this function uses longform
  group_by(LOCATION) %>% 
  mutate(bal_05ya = rollmean(deficit_pc, k = 5, fill = NA),
         bal_07ya = rollmean(deficit_pc, k = 7, fill = NA),
         bal_10ya = rollmean(deficit_pc, k = 10, fill = NA)) %>% 
        ungroup() #this closes that group_by(LOCATION)
```

*deficit_pc and its rolling average recodings: Positive values mean budget surpluses and negative deficits*

## Exploring data

### Plot 1
Visualising balance as % of GDP for all countries over the years (positive values are surpluses, negative values are deficits)
```{r}
  p1 <- df_deficit_pct %>% #making p into a ggplot object to visualise w/ plotly
 # line below would invert surplus/deficit in deficit_pc, making higher values mean higher deficits
 # mutate(deficit_pc=-deficit_pc) %>% 
  ggplot(aes(TIME, deficit_pc, colour=LOCATION)) + 
  geom_line()
  ggplotly(p1) #plotly lets me see the legend when scrolling through the plot
```

### Plot 2
Visualising 5 year rolling averages of balance as % of GDP for all countries 
```{r} 
    p2 <- df_deficit_pct %>%
    ggplot(aes(TIME, bal_05ya, colour=LOCATION)) + 
    geom_line()
    ggplotly(p2)
```

### Plot 3
Visualising 7 year rolling averages of balance as % of GDP for all countries
```{r}
    p3 <- df_deficit_pct %>%
    ggplot(aes(TIME, bal_07ya, colour=LOCATION)) + 
    geom_line()
    ggplotly(p3)
```

### Plot 4
Visualising 10 year rolling averages of balance as % of GDP for all countries
```{r} 
    p4 <- df_deficit_pct %>%
    ggplot(aes(TIME, bal_10ya, colour=LOCATION)) + 
    geom_line()
    ggplotly(p4) 
```


### Plot 5
Visualising 7 year rolling averages of balance as % of GDP for all countries from 2000
```{r}
    p5 <- df_deficit_pct %>%
    filter(TIME>=2000) %>% 
    ggplot(aes(TIME, bal_07ya, colour=LOCATION)) + 
    geom_line()
    ggplotly(p5)
```


### Plot 6
Visualising 7 year rolling averages of deficit % of GDP for all countries 
*higher values mean higher deficit*
```{r}
    p6 <- df_deficit_pct %>%
    ggplot(aes(TIME, -bal_07ya, colour=LOCATION)) + 
    geom_line()
    ggplotly(p6)
```


###Plot 7
Visualising 7 year rolling averages of deficit % of GDP for all countries with highlighted comparators
*higher values mean higher deficit*
```{r}
    p7 <- df_deficit_pct %>%
    ggplot(aes(TIME, -bal_07ya, colour=in_comparators, group=LOCATION)) + 
    geom_line()
    ggplotly(p7)
```


### Plot 8
Visualising deficit as % of GDP for all countries with highlighted comparators
*higher values mean higher deficit*
```{r}
    p8 <- df_deficit_pct %>%
    ggplot(aes(TIME, -deficit_pc, colour=in_comparators, group=LOCATION)) + 
    geom_line()
    ggplotly(p8)
```

## Thoughts

* Using the whole timeline gives us Finland during its economic crisis which might be interesting for the point that countries borrow more during crises
* The shorter timeline (starting at around 2000) would be more parsimonious
* We can choose between 5, 7 or 10 rolling averages. I think 7 might be the fairer picture. 
* Rolling averages make us lose data points for the more recent years where it can't be calculated like that. It would be good to think about whether that's a tradeoff that we're happy with. 
* General story seems to be that no one has run the kind of deficit that Scotland might be looking at consistently and particularly not outside of crisis periods. 
* Interesting that Norway is the country Scotland is most often compared with and it's a total outlier in terms of running surpluses.
* Ireland, another potential comparison, is on the other end of that. 
* This group of comparators I chose ("DNK","IRL","NZL","NOR","SWE") might not be the best one, I picked them because they're some of the comparators that come up more often. Might make more sense to stick to the countries that the Sustainable Growth Commission uses as their case studies. 
* I think the best way to visualise this in the actual paper would probably be this kind of chart (obviously with very different formatting and less crowded) or if we want to just stick to the comparators we could do are bar charts (the ones that have the change graph at their tip)
