---
title: "Long term interest rates - GDP - debt-to-GDP ratio"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library (tidyverse)
library (plotly)
```

## Data cleaning

Uploading data
```{r, message=FALSE, warning=FALSE}
total <- read_csv("data/LTINT_GDP.csv")
df_gdp <- read_csv("data/GDP_yearly.csv")
df_ltint <- read_csv("data/LTINT_yearly.csv")
```

Making data wide
I removed the Flag Codes to make the wide frames more manageable. Flags were not necessary for the analysis.
```{r}
df_ltint_wide <- df_ltint %>% 
  select(-`Flag Codes`) %>% 
  pivot_wider(names_from = INDICATOR, values_from = Value)
```

Using MEASURE variable as the pivot to make it wide (names_from = MEASURE) because there were two different measures of GDP (million USD and per capita)
Then dropping the INDICATOR variable that had GDP as values
```{r}
df_gdp_wide <- df_gdp %>% 
  select(-`Flag Codes`) %>% 
  pivot_wider(names_from = MEASURE, values_from = Value, names_prefix = "GDP_") %>% 
  select(-INDICATOR)
```

Merging GDP and long term interest rates in one dataframe
```{r}
df_merged <- df_gdp_wide %>% left_join(df_ltint_wide %>% select(LOCATION, TIME, LTINT), by = c("LOCATION","TIME"))
```

```{r}
df_debt_ratio <- read_csv("data/debt_gdp_ratio_oecd.csv")

```


```{r}
df_debt_ratio <- df_debt_ratio %>% rename(debt_gdp = Value)
```

```{r}
df_debt_wide <- df_debt_ratio %>% 
  select(-`Flag Codes`) %>% 
  pivot_wider(names_from = MEASURE, values_from = debt_gdp) %>% 
  select(-INDICATOR)
df_debt_wide
```
Merging dataframes (long term interest rates, debt-to-gdp ratio and gdp)
```{r}
df_merged <- df_gdp_wide %>% 
  left_join(df_ltint_wide %>% select(LOCATION, TIME, LTINT), by = c("LOCATION","TIME")) %>% 
  left_join(df_debt_wide %>% select(LOCATION, TIME, PC_GDP), by = c("LOCATION","TIME")) %>% 
  rename(debt_gdp = PC_GDP)
df_merged
```

Creating a natural log of GDP variable
```{r}
df_merged <- df_merged %>%
  mutate(ln_gdp = log(GDP_MLN_USD)) 
``` 

Creating dataset from 2006 (date where long term interest rates data starts)
```{r}
df_merged_2006_2020 <- df_merged %>% filter(TIME>=2006)
df_merged_2006_2020 %>% write_csv("gdp_ltint_debt2006_2020.csv")
```

Comparing what countries are in each dataframe
```{r}
countries_gdp <- df_gdp %>% select(LOCATION) %>% unique()
countries_debt <- df_debt_ratio %>% select(LOCATION) %>% unique()
countries_ltint <- df_ltint %>% select(LOCATION) %>% unique()

setdiff(countries_gdp, countries_debt)
setdiff(countries_debt, countries_ltint)
setdiff(countries_gdp, countries_ltint)

```

```{r}
df_merged_2006_2020 %>% summary(debt_gdp)
median_debt <- median(df_merged_2006_2020$debt_gdp, na.rm=T)
```

Creating a eurozone countries variable
```{r}
eurozone <- c("AUT","BEL","CYP","EST","FIN","FRA","DEU","GRC","IRL","ITA","LVA","LTU","LUX","MLT","NLD","PRT","SVK","SVN","ESP")
```

Creating and above-below variable for median of debt-to-gdp in 2013
```{r}
median_debt_euro_2019 <- df_merged_2006_2020 %>% 
  filter(TIME==2013, LOCATION %in% eurozone) %>% 
  pull(debt_gdp) %>% median(na.rm=T)

```

Creating a comparators variable for those with similar debt-to-gdp ratios
```{r}
comp_debt_gdp <- c("BEL","FRA","DEU","IRL","LUX","MLT","NLD","SVK","ESP")
```

## Data exploring

*Debt-to-gdp ratio data for 2020 is very patchy so 2019 will be used as the most recent data point*

### Plot 1

Visualising the relationship between GDP and long term interest rates in 2019 
Colour of the dot corresponds to whether the country is above or below the median debt-to-gdp ratio in 2019
```{r}
median_debt_2019 <- df_merged_2006_2020 %>% 
  filter(TIME==2019, LOCATION!="EU27_2020", LOCATION!="OECDE", LOCATION!="EA19") %>% 
  pull(debt_gdp) %>% median(na.rm=T)


```    

```{r}
p1 <- df_merged_2006_2020 %>% 
    filter(TIME==2019, LOCATION!="EU27_2020", LOCATION!="OECDE", LOCATION!="EA19") %>% 
    mutate(above_debt_median=if_else(debt_gdp>=median_debt_2019,TRUE,FALSE)) %>% 
    ggplot(aes(LTINT, GDP_MLN_USD, colour=above_debt_median, group=LOCATION)) + 
    geom_point()
ggplotly(p1) 
```    

Visualising the relationship between GDP and long term interest rates in 2013
```{r}
median_debt_2013 <- df_merged_2006_2020 %>% 
  filter(TIME==2013, LOCATION!="EU27_2020", LOCATION!="OECDE", LOCATION!="EA19") %>% 
  pull(debt_gdp) %>% median(na.rm=T)


```    

### Plot 2

Scatterplot - Colour of the dot corresponds to whether the country is above or below the median debt-to-gdp ratio in 2013
```{r}
p2 <- df_merged_2006_2020 %>% 
    filter(TIME==2013, LOCATION!="EU27_2020", LOCATION!="OECDE", LOCATION!="EA19") %>% 
    mutate(above_debt_median=if_else(debt_gdp>=median_debt_2013,TRUE,FALSE)) %>% 
    ggplot(aes(LTINT, GDP_MLN_USD, colour=above_debt_median, group=LOCATION)) + 
    geom_point()
ggplotly(p2) 
```    

### Plot 3

Visualising long term interest rates and GDP relationship, colour-coded by above or below median of eurozone debt-to-gdp ratio
```{r}
p3 <- df_merged_2006_2020 %>% 
    filter(TIME==2019, LOCATION %in% eurozone) %>% 
    mutate(above_debt_median=if_else(debt_gdp>=median_debt_euro_2019,TRUE,FALSE)) %>% 
    ggplot(aes(LTINT, GDP_MLN_USD, colour=above_debt_median, group=LOCATION)) + 
    geom_point()
ggplotly(p3) 
```    

### Plot 4

Visualising long term interest rates and debt-to-gdp ratio relationship in eurozone
```{r}
p4 <- df_merged_2006_2020 %>% 
    filter(TIME==2019, LOCATION %in% eurozone) %>% 
    ggplot(aes(LTINT, debt_gdp, colour=LOCATION)) + 
    geom_point()
ggplotly(p4) 

```    

### Plot 5

Visualising GDP and debt-to-gdp ratio relationship in eurozone to choose a good comparator (similar debt-to-gdp ratio but different sizes)
```{r}
p5 <- df_merged_2006_2020 %>% 
    filter(TIME==2019, LOCATION %in% eurozone) %>% 
    ggplot(aes(GDP_MLN_USD, debt_gdp, colour=LOCATION)) + 
    geom_point()
ggplotly(p5) 
```    


## Thoughts

* We will probably have to use 2019 as the most recent data point throughout because debt-to-gdp data for 2020 is patchy here. 
* We can also do analysis for 2020 with the other variables. 
* We can do everything just for the eurozone to eliminate noise and the effect of the US as an outlier but we might also choose to do EU (including non-eurozone) and UK to get a bit more info but also eliminate less noise
* Would be good to decide whether the UK always being there is necessary, I think it would probably be good. 
* I used the median as the comparison point for debt-to-gdp ratio but we might want to choose something else. 
* 2013 does not look that different from 2019 from what I can tell. 
* To visualise I can't think of much else beyond scatterplots, need to pick whether independent variable we care most about is debt-to-gdp ratio or GDP (or both). Could do two different plots, could do debt-to-gdpXlong term interest rates with GDP as size of the dots. 

### Comparators

* Option 1: Theory-driven comparators. SGC used New Zealand, 
* Option 2: Countries that have similar debt-to-gdp ratios but very different GDPs, to see how they compare in terms of long term interest rates. Countries with similar debt-to-gdp ratios to Germany might make for good comparators with very different GDPs. Maybe the Netherlands, Finland, Ireland, Slovakia. See plot 5.

### Plot 6

Same as plot 3 but with natural log of GDP
Visualising long term interest rates and natural log of GDP relationship, colour-coded by above or below median of eurozone debt-to-gdp ratio
```{r}
p6 <- df_merged_2006_2020 %>% 
    filter(TIME==2019, LOCATION %in% eurozone) %>% 
    mutate(above_debt_median=if_else(debt_gdp>=median_debt_euro_2019,TRUE,FALSE)) %>% 
    ggplot(aes(LTINT, ln_gdp, colour=above_debt_median, group=LOCATION)) + 
    geom_point()
ggplotly(p6) 
```    

### Plot 7

Visualising relationship between debt-to-GDP ratio and long term interest rates with size of bubbles representing the natural log of GDP
```{r}
p7 <- df_merged_2006_2020 %>% 
    filter(TIME==2019, LOCATION %in% eurozone) %>% 
    ggplot(aes(x = LTINT, y = debt_gdp, size = ln_gdp,    colour=LOCATION)) + 
    geom_point(alpha=0.7)
ggplotly(p7) 
```    


### Plot 8

Visualising relationship between natural log of GDP and long term interest rates only using countries with similar debt-to-GDP ratios
```{r}
p8 <- df_merged_2006_2020 %>% 
    filter(TIME==2019, LOCATION %in% comp_debt_gdp) %>% 
    ggplot(aes(x = LTINT, y = ln_gdp, colour=LOCATION)) + 
    geom_point(alpha=0.7)
ggplotly(p8) 
```    