---
---
title: "Data analysis - Deficit as % of GDP"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library (tidyverse)
library (plotly)
library(zoo) # moving averages 
library(kableExtra)
```

## Preparing data

Uploading data 
```{r reading, message=FALSE}
df_deficit_pct <- read_csv("data/deficit_pct_gdp_OECD.csv")
```

Creating a comparators variable
```{r} 
comparators <- c("DNK","IRL","NZL","NOR","SWE")
    df_deficit_pct <- df_deficit_pct %>%
    mutate(in_comparators=if_else(LOCATION%in%comparators,TRUE,FALSE)) 
```

Creating a dataframe that starts in 1995 (when we start having data for all countries)
```{r}
df_deficit_1995on <- df_deficit_pct %>%
  filter(TIME>1995)
```


Renaming deficit_pc to balance of payments
```{r} 
    df_deficit_1995on <- df_deficit_1995on %>%
    mutate(balance_pc = deficit_pc) 
```

Transforming deficit_pc (currently a measure of balance of payments where positive values mean surpluses) into a measure of deficit (higher values mean higher deficits)
```{r} 
    df_deficit_1995on <- df_deficit_1995on %>%
    mutate(deficit_pc = -deficit_pc) 
```

Creating rolling averages over 3 and 5 years
*These variable names will be changed as now they are no longer rolling averages of the balance but of actual deficit*
```{r}
  df_deficit_1995on <- df_deficit_1995on %>% #this function uses longform
  group_by(LOCATION) %>% 
  mutate(def_03ya = rollmean(deficit_pc, k = 3, fill = NA),
         def_05ya = rollmean(deficit_pc, k = 5, fill = NA)) %>% 
        ungroup() #this closes that group_by(LOCATION)
```

**Creating different measures of rolling minimum deficits**
Creating rolling minimums over 3 and 5 years
This is the minimum deficit in rolling 5 year periods including GFC
```{r}
df_deficit_1995on <- df_deficit_1995on %>%
  group_by(LOCATION) %>% 
  mutate(min_05y_all = rollapply(deficit_pc,5,min, fill=NA, align = "right"),
         min_03y_all = rollapply(deficit_pc,3,min, fill=NA, align = "right")
         ) %>% 
  ungroup ()
```

Creating a dataframe that does not take the GFC into account
```{r}
df_deficit_no_GFC <- df_deficit_1995on %>%
  filter(!(TIME%in%c(2009,2010,2011,2012,2013)))
```

Creating rolling minimums over 3 and 5 years
This is the minimum deficit in rolling 5 year periods excluding GFC from 1995
```{r}
df_deficit_no_GFC <- df_deficit_no_GFC %>%
  group_by(LOCATION) %>% 
  mutate(min_05y_no_GFC = rollapply(deficit_pc,5,min, fill=NA, align = "right"),
         min_03y_no_GFC = rollapply(deficit_pc,3,min, fill=NA, align = "right")
         ) %>% 
  ungroup ()
```

Creating a table with maximum values of 5-year rolling minimums with GFC years from 1995
```{r}
max_5ymin <- df_deficit_1995on %>% 
  group_by(LOCATION) %>% 
  arrange(desc(min_05y_all)) %>% 
  slice(1) %>% 
  ungroup()
```

Creating a table with maximum values of 5-year rolling minimums without GFC years from 1995
```{r}
max_5ymin_no_GFC <- df_deficit_no_GFC %>% 
  group_by(LOCATION) %>% 
  arrange(desc(min_05y_no_GFC)) %>% 
  slice(1) %>% 
  ungroup()
```

Creating rolling means of deficit without GFC years

```{r}
  df_deficit_no_GFC <- df_deficit_no_GFC %>% 
  group_by(LOCATION) %>% 
  mutate(def_03ya_no_GFC = rollmean(deficit_pc, k = 3, fill = NA),
         def_05ya_no_GFC = rollmean(deficit_pc, k = 5, fill = NA)) %>% 
        ungroup() 
```


Creating a table with maximum values of 5-year rolling average without GFC years from 1995
```{r}
max_5ya_no_GFC <- df_deficit_no_GFC %>% 
  group_by(LOCATION) %>% 
  arrange(desc(def_05ya_no_GFC)) %>% 
  slice(1) %>% 
  ungroup()
```

Creating a table with maximum values of 5-year rolling average with all years (including GFC) from 1995
```{r}
max_5ya_all <- df_deficit_1995on %>% 
  group_by(LOCATION) %>% 
  arrange(desc(def_05ya)) %>% 
  slice(1) %>% 
  ungroup()
```


## Exploring data

*deficit_pc and its rolling average recodings: Positive values mean budget deficits and negative values mean budget surpluses*

### Plot 1
Visualising deficit as % of GDP for all countries in all years
```{r}
  p1 <- df_deficit_pct %>%
  ggplot(aes(TIME, deficit_pc, colour=LOCATION)) + 
  geom_line()
  ggplotly(p1)
```

### Plot 2
Maximum 5-year minimums for all years (including GFC)
```{r}
    p2 <- max_5ymin %>%
    ggplot(aes(LOCATION,min_05y_all, group=TIME)) + 
    geom_col() +
    coord_flip()
    ggplotly(p2)
```
Interpretation: 

* e.g. Australia: 2.89 is the minimum value recorded between 2008 and 2012. It was recorded in 2012. In 2008, the deficit rose significantly (from -0.72 to 3.75), and started going back down hence the lowest deficit in the 2008 to 2012 5-year period was in 2012 (2.89) and given the high deficits of this period it was the highest minimum deficit that Australia recorded.

* The year that appears when you hover is the year in which the value was first considered the minimum 5-year value, not the year in which this deficit occurred 

* e.g. Germany (see table below): 2005, 3.03. This value occurred in 2001, and in the 5-year period between 2001 and 2005. In 2001, there was a sharp increase in deficit (from 1.58 to 3.03), and the deficit kept growing. Until 2004, the minimum 5-year deficit was from 2000, before this sharp increase. In 2005, the minimum 5-year value is the one for 2001, where the sharp increase happened. 


```{r} 
tab_de <- df_deficit_1995on %>%  
  select(LOCATION, TIME, deficit_pc, min_05y_all) %>%
  filter(LOCATION == "DEU" & TIME %in% c(1999,2000,2001,2002,2003,2004,2005,2006))

  kbl(x = tab_de) %>%
  kable_paper(bootstrap_options = "striped", full_width = F, position = "left")
```

* For countries that ran surpluses, the value is the maximum minimum deficit, which means the maximum surplus it ran in a 5-year period. 

### Plot 3

Maximum 5-year minimums excluding GFC
```{r}
    p3 <- max_5ymin_no_GFC %>%
    ggplot(aes(LOCATION,min_05y_no_GFC, group=TIME)) + 
    geom_col() +
    coord_flip()
    ggplotly(p3)
```


### Plot 4

Maximum 5-year averages for all years (including GFC)
```{r}
    p4 <- max_5ya_all %>%
    ggplot(aes(LOCATION,def_05ya, group=TIME)) + 
    geom_col() +
    coord_flip()
    ggplotly(p4)
```


### Plot 5

Maximum 5-year averages excluding GFC
```{r}
    p5 <- max_5ya_no_GFC %>%
    ggplot(aes(LOCATION,def_05ya_no_GFC, group=TIME)) + 
    geom_col() +
    coord_flip()
    ggplotly(p5)
```

Exporting data to CSV

```{r}
write_csv(df_deficit_1995on,"deficit_1995on_data.csv")
write_csv(df_deficit_no_GFC,"deficit_noGFC_data.csv")
write_csv(max_5ya_all,"max_5ya_total_table.csv")
write_csv(max_5ya_no_GFC,"max_5ya_noGFC_table.csv")
write_csv(max_5ymin,"max_5ymin_total_table.csv")
write_csv(max_5ymin_no_GFC,"max_5ymin_noGFC_table.csv")
```
